\documentclass[10pt]{article}

\usepackage[margin=0.7in,top=0.6in,bottom=0.6in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\usepackage{microtype}
\setlength{\emergencystretch}{2em}
\sloppy
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\pagestyle{empty}

\newcommand{\code}[1]{\texttt{\small #1}}

\begin{document}

{\Large\bfseries Pen-DHRL-TD: Teacher-Distilled Hierarchical RL for Autonomous Penetration Testing}\\[2pt]
{\small NASimEmu-enhanced Project \hfill \today}

\vspace{2pt}
\hrule
\vspace{6pt}

\textbf{Summary.} Pen-DHRL-TD augments a deep hierarchical RL pentesting agent
(\code{NASimNetDHRL}: a recurrent GNN \emph{manager} + goal-conditioned graph-attention
\emph{worker}, 212k params) with offline supervision from a locally-hosted LLM teacher
(\code{qwen2.5:3b-instruct} via Ollama, never queried inside the training loop). The
manager selects one of 8 named semantic subgoals (e.g.\ \code{GAIN\_INITIAL\_ACCESS},
\code{ESCALATE\_PRIVILEGE}); the teacher labels sanitized, partially-observable state
summaries with a goal + confidence, producing a provenance-tracked dataset distilled
into the subgoal head via a confidence-weighted cross-entropy loss, folded into a
four-stage schedule (supervised warm-start~$\to$~decaying joint PPO+distillation~$\to$~guaranteed
RL-only tail). A live \emph{explainability monitor} prints the manager's realized
subgoal histogram and distillation diagnostics every epoch.

\textbf{Method highlights.}
\begin{itemize}[nosep,leftmargin=1.2em]
  \item \textbf{Teacher pipeline:} discovered-only state summary $\to$ JSON-schema-constrained
    LLM query $\to$ anti-hallucination/executable-text validator $\to$ retry-on-reject,
    never silently discarded.
  \item \textbf{Triggered sampling:} teacher queried only at informative moments
    (new subnet, first compromise, privilege change, IDS alert, repeated failure, high
    manager entropy, periodic fallback) under a diversified rollout-policy mixture.
  \item \textbf{Distillation loss:} $\mathcal{L}_{\text{goal}} = \text{mean}_i\big[
    \operatorname{clip}(c_i, 0.05, 1)\cdot(-\log \hat p_\theta(g_i\mid s_i))\big]$,
    jointly trained with PPO under an annealed weight, forced to exactly 0 for the
    final 10\% of training.
\end{itemize}

\textbf{Key findings} (12-subnet, 72--96-host dynamic scenario family; run \code{b33undf9}, 200 epochs):
\begin{itemize}[nosep,leftmargin=1.2em]
  \item \textbf{Teacher dataset:} 2{,}235 labeled states, 89.5\% valid, but severely
    class-imbalanced (84\% \code{DISCOVER\_SUBNET}; \code{RECOVER\_OR\_REPLAN}: 0 examples),
    which the live subgoal histogram traced directly to a training-time subgoal
    collapse (92\%$\to$100\% \code{DISCOVER\_SUBNET} by training's end).
  \item \textbf{Checkpoint-selection pitfall:} the default ``best-checkpoint'' metric
    (\code{captured\_avg}) is blind to episode length and froze on a mid-run checkpoint,
    missing a large late-training efficiency gain visible only in \code{reward\_avg\_episodes}.
  \item \textbf{Outperforms a non-distilled Pen-DHRL baseline:} same architecture,
    same parameter count (212k), trained \emph{without} the teacher signal. The
    distilled agent wins on every metric, every scenario.
\end{itemize}

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrrr}
\toprule
Scenario & reward\_avg & reward\_avg\_episodes & eplen\_avg & captured\_avg \\
\midrule
\multicolumn{5}{l}{\textit{Pen-DHRL-TD (final checkpoint, LLM-distilled)}} \\
base   & 2.3216 & $\mathbf{259.12}$ & $\mathbf{111.61}$ & $\mathbf{29.63}$ \\
varA   & 0.2967 & $\mathbf{80.47}$  & $\mathbf{271.18}$ & $\mathbf{17.50}$ \\
varB   & 0.1688 & $\mathbf{46.04}$  & $\mathbf{272.69}$ & $\mathbf{13.25}$ \\
bridge & 1.7425 & $\mathbf{205.48}$ & $\mathbf{117.93}$ & $\mathbf{24.67}$ \\
\multicolumn{5}{l}{\textit{Pen-DHRL baseline (same architecture/params, no distillation)}} \\
base   & 0.5428 & 217.11 & 400.0 & 29.59 \\
varA   & 0.1708 & 68.33  & 400.0 & 13.73 \\
varB   & 0.0778 & 31.12  & 400.0 & 9.39 \\
bridge & 1.1470 & 161.96 & 141.20 & 23.36 \\
\bottomrule
\end{tabular}
\caption{\small 100 held-out episodes/scenario. Baseline: stronger of two
independently trained runs per scenario.}
\end{table}

\textbf{Takeaway.} Pen-DHRL-TD wins on every metric on every scenario, and
terminates well before the 400-step limit on all four (the baseline only does so on
\code{bridge}). Because the two runs share architecture and parameter count exactly,
the gain traces to the LLM-teacher distillation signal itself, not extra model
capacity.

{\small\textit{Full technical report (18 pp.: architecture, teacher pipeline,
5 algorithms, complete results and limitations) available on request.}}

\end{document}
